---
title: "Step-by-step guide to the HDcox pipeline"
author: 
    name: "Pedro Salguero García"
    affiliation: "Institute for Integrative Systems Biology (I2SysBio) and Polytechnic University of Valencia, Valencia, Spain"
    email: pedrosalguerog@gmail.com
package: HDcox
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Step-by-step guide to the HDcox pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
.main-container {
  max-width: 1080px !important;
  margin-left: auto;
  margin-right: auto;
}

body {
  max-width: 1080px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r, include = FALSE}
dpi = 125

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=dpi, 
  fig.retina=1, 
  fig.width=1440/dpi, #4:3 FHD
  fig.height=1080/dpi, 
  out.width="100%",
  crop = NULL,
  warning = T, 
  error = T
)

rm(dpi)
```

# Introduction

The **HDcox** R package contains the necessary functions to reproduce the
pipeline in [this paper*](),
a study by *Salguero-Garcia et al.* in which we analyze ....

The pipeline includes ... basic analysis blocks:

1. **Cross-validation and performing survival models for high dimensional data sets**. First, ... 

2. **Comparing multiple survival models**. ...

3. **Understanding the results in terms of the original variables**. ...

# Installation

HDcox can be installed from GitHub using `devtools`:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("ConesaLab/HDcox", build_vignettes = TRUE)
```

# Getting ready

To run the analyses in this vignette, you'll first need to load `HDcox`:

```{r setup, results = "hide"}
# load HDcox
library(HDcox)
```

In addition, we'll require some additional packages for data formatting. Most of them are signaled as `HDcox` dependencies, 
so they will already be installed in your system.

To generate plots, we make use of the `RColorConesa` R package.
After install:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ConesaLab/RColorConesa")
```

...you can load and set the theme of your R session as follows:

```{r}
library(RColorConesa)
#theme_set(theme_colorConesa()) #under development
```

# Input data

The HDcox pipeline requires two matrices as input. First one must be the features under study, and the second one a matrix with two columns called **time** and **event** for survival information.

To generate the matrices, we first processed the data from ... Details to this process can be found in [our manuscript]() 
(see Supplementary Note and Methods).

If you wish to reproduce the analyses in Salguero-Garcia et al. [[1]](#1)), you can load the
`data_E.MTAB.386.RData` object to use our dataset:

```{r}
# load Tasic dataset
data("X_small_data_E.MTAB.386")
data("Y_small_data_E.MTAB.386")

X <- X_small_data_E.MTAB.386
Y <- Y_small_data_E.MTAB.386

rm(X_small_data_E.MTAB.386, Y_small_data_E.MTAB.386)
```

These contain two `data.frame` objects. After quality control
(see Methods in [Salguero-Garcia et al.]() 
[[1]](#1)), the `data_E.MTAB.386.RData` data contains expression data for **126 observations** 
and **10357 genes**:

```{r, echo = FALSE}
knitr::kable(X[1:5,1:5])

knitr::kable(Y[1:5,])
```

```{r}
ggp_density.event <- plot_events(Y = Y, roundTo = 0.25, categories = c("Censored","Death")) #name for F and T
```

```{r fig.small = T}
ggp_density.event$plot
```

# Cross-validation and performing survival models for high dimensional data sets

...

HDcox provides multiple functions with different methodologies.

* Classical approaches: cox, coxSW and coxEN
* PLS approaces: plsicox, splsdrcox, splsdrcox_mixOmics and plsdacox_mixOmics

More information for each approach ...

And also, we are going to split the data into train/test (70-30 %).

## TRAIN - TEST

```{r}
set.seed(123)
index_train <- caret::createDataPartition(Y$event,
                                          p = .7, #70% train
                                          list = FALSE,
                                          times = 1)

X_train <- X[index_train,] #1103
Y_train <- Y[index_train,]
X_test <- X[-index_train,] #472
Y_test <- Y[-index_train,]
```

## Clasical approach - !!!

If we tried classical approaches...

```{r, eval = FALSE, message=T, error=F}
# classical approach
cox_model <- cox(X = X_train, Y = Y_train, x.center = T, x.scale = F)
```

Classical approaches can not manage high dimensional data and the assumption of having a value greater than 5/10 for the **EPV** (events per variable) is not reached

```{r}
EPV <- sum(Y_train$event==1) / ncol(X_train)
EPV
```

So we are ready to run cross-validation techniques in order to select a low number of variables and/or reduce the dimensionality of the data with the PLS methods.

## PLS-ICOX

Also we are going to run cross-validation techniques for PLS-ICOX method.

```{r, eval = FALSE, message=F}
# run cv.plsicox
cv.plsicox_res <- cv.plsicox(X = X_train, Y = Y_train,
                           max.ncomp =  4,
                           n_run = 2, k_folds = 10, 
                           x.scale = T,
                           remove_near_zero_variance = F, remove_zero_variance = F,
                           PARALLEL = T, verbose = F)
cv.plsicox_res #1min 8s.
```

```{r, eval = FALSE, fig.small=T}
# plot cv.plsicox
cv.plsicox_res$plot_AUC
```

So plsicox_model ...

```{r}
plsicox_model <- plsicox(X = X_train, Y = Y_train, 
                         n.comp = 3, #n.comp = cv.plsicox_res$opt.comp
                         x.center = T, x.scale = F)

plsicox_model
```

## sPLS-DRCOX

Also we are going to run cross-validation techniques for sPLS-DRCOX method.

```{r, eval = FALSE, message=F}
# run cv.splsdrcox
cv.splsdrcox_res <- cv.splsdrcox(X = X_train, Y = Y_train, 
                                 max.ncomp = 4, eta.list = seq(0,0.5,0.25), #penalty
                                 n_run = 2, k_folds = 10, 
                                 x.scale = T, 
                                 remove_near_zero_variance = F, remove_zero_variance = F,
                                 PARALLEL = T, verbose = F)

cv.splsdrcox_res #2min 40s.
```

So splsdrcox_model ...

```{r}
splsdrcox_model <- splsdrcox(X = X_train, Y = Y_train, 
                             n.comp = 2, eta = 0.25, #n.comp = cv.splsdrcox_res$opt.comp, eta = cv.splsdrcox_res$opt.eta
                             x.center = T, x.scale = F)

splsdrcox_model
```

## PLS-DACOX

Also we are going to run cross-validation techniques for PLSDACOX of the PLS methods.

```{r, eval = FALSE, message=F}
# run cv.splsdrcox
cv.plsdacox_res <- cv.plsdacox_mixOmics(X = X_train, Y = Y_train, 
                                        max.ncomp = 4,  #penalty
                                        n_run = 2, k_folds = 10, 
                                        x.scale = T, 
                                        remove_near_zero_variance = F, remove_zero_variance = F,
                                        PARALLEL = T, verbose = F)

cv.plsdacox_res #2min
```

So plsdacox_model ...

```{r}
plsdacox_model <- plsdacox_mixOmics(X = X_train, Y = Y_train, 
                                    n.comp = 3, #cv.plsdacox_res$opt.comp
                                    x.center = T, x.scale = F)

plsdacox_model
```

# Comparing multiple survival models

## Comparing for multiple evaluators at the same time.

```{r}
lst_models <- list("PLS-ICOX" = plsicox_model,
                   "SPLS-DRCOX" = splsdrcox_model,
                   "PLS-DACOX" = plsdacox_model)

eval_results <- eval_models4.0(lst_models = lst_models,
                               X_test = X_test, Y_test = Y_test, 
                               pred.method = "cenROC",
                               pred.attr = "mean",
                               times = seq(1,4,0.5), max_time_points = 15, 
                               PARALLEL = T)

# lst_evaluators <- c(cenROC = "cenROC", 
#                     risksetROC = "risksetROC")
# 
# eval_results <- purrr::map(lst_evaluators, ~eval_models4.0(lst_models = lst_models,
#                                                            X_test = X_test, Y_test = Y_test, 
#                                                            pred.method = .,
#                                                            pred.attr = "mean",
#                                                            times = seq(1,4,0.5), max_time_points = 15, 
#                                                            PARALLEL = T))
```

```{r}
eval_results
#eval_results$cenROC
```

## Plot comparison

```{r}
lst_eval_results <- plot_evaluation(eval_results)
#lst_eval_results <- plot_evaluation.list(eval_results)
```

```{r, fig.small=T}
lst_eval_results$lst_plots$lineplot.mean
lst_eval_results$lst_plot_comparisons$t.test

# lst_eval_results$cenROC$lst_plots$lineplot.mean
# lst_eval_results$cenROC$lst_plot_comparisons$t.test
```

## Computing time comparison (we use the cross validation models)

We should add the cross validations models in order to see which method takes more time.

```{r}
lst_models_time <- list(plsicox_model,
                        splsdrcox_model,
                        plsdacox_model,
                        eval_results)
```

```{r, fig.small=T}
ggp_time <- plot_time.models(lst_models_time)
ggp_time
```

## Forest plots

```{r}
lst_forest_plot <- purrr::map(lst_models, ~survminer::ggforest(.$survival_model$fit, 
                                                               data = .$survival_model$fit$model))
```

```{r, fig.small=T}
lst_forest_plot$`SPLS-DRCOX`
```

## Density plots for each prediction

```{r}
density.plots.lp <- plot_cox.event.list(lst_models, type = "lp")
```

```{r, fig.small=T}
density.plots.lp$`SPLS-DRCOX`$plot.density
density.plots.lp$`SPLS-DRCOX`$plot.histogram
```

## PH Assumption

```{r}
lst_ph_ggplot <- plot_proportionalHazard.list(lst_models)
```

```{r, fig.small=T}
lst_ph_ggplot$`SPLS-DRCOX`
```

# Understanding the results in terms of the original variables

## Psudobeta

```{r}
ggp.simulated_beta <- plot_pseudobeta.list(lst_models = lst_models, 
                                           error.bar = T, onlySig = T, alpha = 0.05, 
                                           zero.rm = T, auto.limits = T, top = 20)
```

```{r, fig.small=T}
ggp.simulated_beta$`SPLS-DRCOX`$plot
```

## Kaplan-Meier

### Full model

```{r}
LST_KM_RES_LP <- getAutoKM.list(type = "LP",
                                lst_models = lst_models,
                                comp = 1:4,
                                top = 10,
                                ori_data = T,
                                BREAKTIME = NULL,
                                only_sig = T, alpha = 0.05)
```

```{r, fig.small=T}
LST_KM_RES_LP$`SPLS-DRCOX`$LST_PLOTS$LP
```

### Per components

```{r}
LST_KM_RES_COMP <- getAutoKM.list(type = "COMP",
                                  lst_models = lst_models,
                                  comp = 1:4,
                                  top = 10,
                                  ori_data = T,
                                  BREAKTIME = NULL,
                                  only_sig = T, alpha = 0.05)
```

```{r, fig.small=T}
LST_KM_RES_COMP$`SPLS-DRCOX`$LST_PLOTS$comp_1
LST_KM_RES_COMP$`SPLS-DRCOX`$LST_PLOTS$comp_2
```

### Per original variables

Matches psudobetas

```{r}
LST_KM_RES_VAR <- getAutoKM.list(type = "VAR",
                                 lst_models = lst_models,
                                 comp = 1:4,
                                 top = 10,
                                 ori_data = T,
                                 BREAKTIME = NULL,
                                 only_sig = T, alpha = 0.05)
```

```{r, fig.small=T}
LST_KM_RES_VAR$`SPLS-DRCOX`$LST_PLOTS$POSTN
LST_KM_RES_VAR$`SPLS-DRCOX`$LST_PLOTS$SIRT5
```

# New patients

```{r}
new_pat <- X_test[1,,drop=F]
```

```{r}
knitr::kable(Y_test[rownames(new_pat),])
```

```{r}
ggp.simulated_beta_newPat <- plot_pseudobeta.newPatient.list(lst_models = lst_models, 
                                                             new_pat = new_pat,
                                                             error.bar = T, onlySig = T, alpha = 0.05,
                                                             zero.rm = T, auto.limits = T, show.betas = T, top = 20)
```

Opposite direction means its value is under the mean for risk one or over the mean for protective ones.

```{r, fig.small=T}
ggp.simulated_beta_newPat$`SPLS-DRCOX`$plot
```

## Add patient to density plot

```{r}
pat_density <- plot_patient.eventDensity(patient = new_pat, 
                                         time = NULL, 
                                         model = lst_models$`SPLS-DRCOX`, 
                                         type = "lp")
```

```{r, fig.small=T}
pat_density
```

```{r}
pat_histogram <- plot_patient.eventHistogram(patient = new_pat, 
                                             time = NULL, 
                                             model = lst_models$`SPLS-DRCOX`, 
                                             type = "lp")
```

```{r, fig.small=T}
#plot_divergent.biplot - for num and qual variables
```

## COX compare patients

```{r}
lst_cox.comparison <- plot_cox.comparePatients.list(lst_models = lst_models, 
                                     df.pat = X_test[1:5,], 
                                     error.bar = T, zero.rm = T, onlySig = T, alpha = 0.05, top = 5)
```

```{r}
knitr::kable(Y_test[1:5,])
```

```{r, fig.small=T}
lst_cox.comparison$`SPLS-DRCOX`$plot
```
