---
title: "Step-by-step guide to the HDcox pipeline"
author: 
  - name: "Pedro Salguero García"
    affiliation: "Institute for Integrative Systems Biology (I2SysBio) and Polytechnic University of Valencia, Valencia, Spain"
    email: pedrosalguerog@gmail.com
package: HDcox
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Step-by-step guide to the HDcox pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 150, 
  out.width="100%",
  crop = NULL
)
```

# Introduction

The **HDcox** R package contains the necessary functions to reproduce the
pipeline in [this paper*](),
a study by *Salguero-Garcia et al.* in which we analyze ....

The pipeline includes ... basic analysis blocks:

1. **Cross-validation and performing survival models for high dimensional data sets**. First, ... 

2. **Comparing multiple survival models**. ...

3. **Understanding the results in terms of the original variables**. ...

# Installation

HDcox can be installed from GitHub using `devtools`:

```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("ConesaLab/HDcox", build_vignettes = TRUE)

```

# Getting ready

To run the analyses in this vignette, you'll first need to load `HDcox`:

```{r setup, results = "hide"}
# load HDcox
library(HDcox)

# load auxiliary packages
suppressPackageStartupMessages({
  library(ggforce)
  library(grDevices)
  library(splines)
  library(RColorConesa)
})
```

In addition, we'll require some additional packages for data formatting. Most of them are signaled as `HDcox` dependencies, 
so they will already be installed in your system.

To generate plots, we make use of the `RColorConesa` R package.
After install:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("ConesaLab/RColorConesa")
```

...you can load and set the theme of your R session as follows:

```{r}
library(RColorConesa)
#theme_set(theme_colorConesa()) #under development
```

# Input data

The HDcox pipeline requires two matrices as input. First one must be the features under study, and the second one a matrix with two columns called **time** and **event** for survival information.

To generate the matrices, we first processed the data from ... Details to this process can be found in [our manuscript]() 
(see Supplementary Note and Methods).

If you wish to reproduce the analyses in Salguero-Garcia et al. [[1]](#1)), you can load the
`data_E.MTAB.386.RData` object to use our dataset:

```{r}
# load Tasic dataset
data("data_E.MTAB.386")
```

These contain two `data.frame` objects. After quality control
(see Methods in [Salguero-Garcia et al.]() 
[[1]](#1)), the `data_E.MTAB.386.RData` data contains expression data for **126 observations** 
and **10357 genes**:

```{r, echo = FALSE}
knitr::kable(X[1:5,1:5])

knitr::kable(Y[1:5,])
```

# Cross-validation and performing survival models for high dimensional data sets

...

HDcox provides multiple functions with different methodologies.

* Classical approaches: cox, coxSW and coxEN
* PLS approaces: plsicox, splsdrcox, splsdrcox_mixOmics and plsdacox_mixOmics

More information for each approach ...

To speed-up the analysis we are going to reduce the number of variables of X.

```{r, eval = FALSE}
set.seed(123)
X <- X[,sample(1:ncol(X), 500)]
```

And also, we are going to split the data into train/test (70-30 %).

```{r, eval = FALSE}
set.seed(123)
index_train <- caret::createDataPartition(Y$event,
                                          p = .7, #70% train
                                          list = FALSE,
                                          times = 1)

X_train <- X[index_train,] #1103
Y_train <- Y[index_train,]
X_test <- X[-index_train,] #472
Y_test <- Y[-index_train,]
```

If we tried classical approaches...

```{r, message=T, error=F}
# classical approach
cox_model <- cox(X = X_train, Y = Y_train, x.center = T, x.scale = F)
```

Classical approaches can not manage high dimensional data and the assumption of having a value greater than 5/10 for the **EPV** (events per variable) is not reached

```{r, eval = FALSE}
EPV <- sum(Y_train$event==1) / ncol(X_train)
EPV
```

So we are ready to run cross-validation techniques in order to select a low number of variables (coxEN) or reduce the dimensionality of the data with the PLS methods.

```{r, eval = FALSE, message=F, warning=F}
# run cv.coxEN
cv.coxEN_res <- cv.coxEN(X = X_train, Y = Y_train, 
                         EN.alpha.list = seq(0,1,0.1), #EN penalty
                         n_run = 2, k_folds = 10, 
                         x.scale = T,
                         remove_near_zero_variance = F, remove_zero_variance = F,
                         PARALLEL = T, verbose = F)
cv.coxEN_res #1min 10s.
```

So coxEN_model ...

```{r}
coxEN_model <- coxEN(X = X_train, Y = Y_train, EN.alpha = 0,
                     x.center = T, x.scale = F)

coxEN_model
```

Also we are going to run cross-validation techniques for one of the PLS methods.

```{r, eval = FALSE, message=F, warning=F}
# run cv.splsdrcox
cv.splsdrcox_res <- cv.splsdrcox(X = X_train, Y = Y_train, 
                                 max.ncomp = 4, eta.list = c(0,0.25,0.5,0.75), #penalty
                                 n_run = 2, k_folds = 10, 
                                 x.scale = T, 
                                 remove_near_zero_variance = F, remove_zero_variance = F,
                                 PARALLEL = T, verbose = F)

cv.splsdrcox_res #2min 30s.
```

So splsdrcox_model ...

```{r}
splsdrcox_model <- splsdrcox(X = X_train, Y = Y_train, 
                             n.comp = 2, eta = 0.25,
                             x.center = T, x.scale = F)

splsdrcox_model
```

Also we are going to run cross-validation techniques for one of the PLS methods.

```{r, eval = FALSE, message=F, warning=F}
# run cv.splsdrcox
cv.plsdacox_res <- cv.plsdacox_mixOmics(X = X_train, Y = Y_train, 
                                        max.ncomp = 4,  #penalty
                                        n_run = 2, k_folds = 10, 
                                        x.scale = T, 
                                        remove_near_zero_variance = F, remove_zero_variance = F,
                                        PARALLEL = T, verbose = F)

cv.plsdacox_res #1min 35s
```

So plsdacox_model ...

```{r}
plsdacox_model <- plsdacox_mixOmics(X = X_train, Y = Y_train, 
                                    n.comp = 3, 
                                    x.center = T, x.scale = F)

plsdacox_model
```

# Comparing multiple survival models

Comparing for multiple evaluators at the same time.

```{r}
#lst_evaluations <- c("survivalROC", "cenROC", "nsROC", "smoothROCtime_C", "smoothROCtime_I", "risksetROC")
lst_evaluations <- c("cenROC", "risksetROC")
names(lst_evaluations) <- lst_evaluations

lst_models <- list("coxEN" = coxEN_model,
                   "splsdrcox_mixOmics" = splsdrcox_model,
                   "plsdacox_mixOmics" = plsdacox_model)

eval_results <- purrr::map(lst_evaluations, ~eval_models4.0(lst_models = lst_models,
                                                            X_test = X_test, Y_test = Y_test, 
                                                            pred.method = .,
                                                            pred.attr = "mean",
                                                            times = seq(1,4,0.5), max_time_points = 15, 
                                                            PARALLEL = T))

eval_results$cenROC
```

Plot comparison

```{r}
lst_ggp <- plot.evaluation.list(eval_results)
```
