---
title: "Step-by-step guide to the HDcox pipeline"
author: 
    name: "Pedro Salguero García"
    affiliation: "Institute for Integrative Systems Biology (I2SysBio) and Polytechnic University of Valencia, Valencia, Spain"
    email: pedrosalguerog@gmail.com
package: HDcox
date: "Last updated: `r format(Sys.time(), '%d %B, %Y')`"
output: 
    rmarkdown::html_vignette:
        toc: true
vignette: >
  %\VignetteIndexEntry{Step-by-step guide to the HDcox pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<style type="text/css">
.main-container {
  max-width: 1080px !important;
  margin-left: auto;
  margin-right: auto;
}

body {
  max-width: 1080px !important;
  margin-left: auto;
  margin-right: auto;
}
</style>

```{r, include = FALSE}
dpi = 125

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=dpi, 
  fig.retina=1, 
  fig.width=1440/dpi, #4:3 FHD
  fig.height=1080/dpi, 
  out.width="100%",
  crop = NULL,
  warning = T, 
  error = T
)

rm(dpi)
```

# Introduction

The **HDcox** R package contains the necessary functions to reproduce the
pipeline in [this paper*](),
a study by *Salguero-Garcia et al.* in which we analyze ....

The pipeline includes ... basic analysis blocks:

1. **Cross-validation and performing survival models for high dimensional data sets**. First, ... 

2. **Comparing multiple survival models**. ...

3. **Understanding the results in terms of the original variables**. ...

# Installation

HDcox can be installed from GitHub using `devtools`:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("ConesaLab/HDcox", build_vignettes = TRUE)
```

# Getting ready

To run the analyses in this vignette, you'll first need to load `HDcox`:

```{r setup, eval=FALSE, results = "hide"}
# load HDcox
library(HDcox)
```

In addition, we'll require some additional packages for data formatting. Most of them are signaled as `HDcox` dependencies, 
so they will already be installed in your system.

To generate plots, we make use of the `RColorConesa` R package.
After install:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("ConesaLab/RColorConesa")
```

...you can load and set the theme of your R session as follows:

```{r}
library(RColorConesa)
#theme_set(theme_colorConesa()) #under development
```

# Input data

The HDcox pipeline requires two matrices as input. First one must be the features under study, and the second one a matrix with two columns called **time** and **event** for survival information.

To generate the matrices, we first processed the data from ... Details to this process can be found in [our manuscript]() 
(see Supplementary Note and Methods).

If you wish to reproduce the analyses in Salguero-Garcia et al. [[1]](#1)), you can load the
`data_E.MTAB.386.RData` object. After quality control (see Methods in [Salguero-Garcia et al.]() 
[[1]](#1))the data contains expression data for **552 observations** 
and **534 miRNA**. The data could be load as follows:

```{r}
# load Tasic dataset
data("X_miRNA_glioblastoma")
data("Y_miRNA_glioblastoma")

X <- X_miRNA_glioblastoma
Y <- Y_miRNA_glioblastoma

rm(X_miRNA_glioblastoma, Y_miRNA_glioblastoma)
```

**X** and **Y** are two `data.frame` objects. **X** is related to the explanatory variables. Rows are observations and columns are the different variables of interest. For **Y** matrix, rows are the observations with the same row names as **X**, and it has to have two columns. The first one "time" and the second one "event". "time" refers to the following up time until the event or until the last control if the observation is a right censored observation. The event could be TRUE/FALSE or 1/0 for those observations that have reach the event or are censored. An example could be show in the next code:

```{r, echo = FALSE}
knitr::kable(X[1:5,1:5])

knitr::kable(Y[1:5,])
```
As we said, the dimension of each object is 552x534 for **X** matrix, and 552x2 for **Y** matrix.

```{r, echo = FALSE}
knitr::kable(dim(X), col.names = "X")
knitr::kable(dim(Y), col.names = "Y")
```

**HDcox** has a series of plots to understand the distribution of the data. One of them is the plot of events along time. The function requires the Y matrix. As optional arguments, the user can specify: 

1. *max.breaks* Number of breaks for the histogram. 
2. *roundTo* the minimum numeric value for round break-times values. E.g. 0.25, a value of 1.32 will be rounded to 1.25 (numbers shall be rounded off to multipliers of 0.25).
1. *categories* the name for each category (character vector of length two)
2. *y.text* the name for the y axis

```{r}
ggp_density.event <- plot_events(Y = Y, 
                                 categories = c("Censored","Death"), #name for FALSE/0 (Censored) and TRUE/1 (Event)
                                 y.text = "Number of observations", 
                                 roundTo = 0.5, 
                                 max.breaks = 15) 
```

```{r fig.small = T}
ggp_density.event$plot
```

# Survival models for low/high dimensional data sets

After loading the data, it may be of interest for the user to perform a survival analysis in order to examine the relationship between explanatory variables and the outcome. However, traditional methods are only applicable for low-dimensional datasets. To address this issue, we have developed a set of functions that utilize (s)PLS techniques in combination with Cox analysis for the analysis of high-dimensional datasets.

*HDcox* provides the following methodologies:

* Classical approaches: COX, COX StepWise and COX Elastic Net.
* PLS approaches: PLS-ICOX, sPLS-DRCOX, sPLS-DRCOX-MixOmics and sPLS-DACOX-MixOmics.

More information for each approach could be found in the help section for each function. The function name for each methodology are:

* Classical approaches: cox, coxSW and coxEN.
* PLS approaches: plsicox, splsdrcox, splsdrcox_mixOmics and splsdacox_mixOmics.

To perform a survival analysis with our example, we will use methodologies that can work with high-dimensional data. These are the set of methodologies that use PLS and COX Elastic Net techniques.

The first thing we are going to do is split our data into a training set and a test set. This split will be made with a proportion of 70% of the data for training and 30% for test.

## TRAIN/TEST

First, we will divide our data into a training and testing set using the `createDataPartition` function from the R package caret. We will use a 70%/30% split for training and testing, respectively, and set a seed for reproducible results.

```{r}
set.seed(321)
index_train <- caret::createDataPartition(Y$event,
                                          p = .8, # 80% train
                                          list = FALSE,
                                          times = 1)
```

```{r}
X_train <- X[index_train,] #443x534
Y_train <- Y[index_train,]
X_test <- X[-index_train,] #109x534
Y_test <- Y[-index_train,]
```

## Clasical approach

We have already mentioned that we need to work with a high-dimensional methodology. However, if we try to run a classical analysis with one of our functions such as a Cox analysis using the entire matrix X, we will get an error due to the specification of the MIN_EPV parameter. This parameter, set by default to 5, establishes a minimum ratio of variables that should be included in a survival model according to the number of patients who experience the event (Events / Variables of X). According to literature, values greater than 10 should be used to obtain models with good predictions, but this value should not be lower than 5.

```{r, eval=FALSE, message=T, error=F}
# classical approach
cox_model <- cox(X = X_train, Y = Y_train, 
                 x.center = T, x.scale = F, 
                 y.center = F, y.scale = F, 
                 remove_near_zero_variance = T, remove_zero_variance = T, toKeep.zv = NULL, 
                 remove_non_significant = F, alpha = 0.05, 
                 MIN_EPV = 5, FORCE = F, returnData = T, verbose = F)
```

Despite specifying an EPV of 0.62, we should have a maximum of 55 variables in our final survival model in relation to the number of events found in our training set. To compute our EPV, we can used the function `getEPV()` that requires the matrix X and Y as input.

```{r}
EPV <- getEPV(X_train, Y_train)
```

```{r}
EPV
```

As previously mentioned, our data set is high dimensional in terms of variables and patients, but also has an EPV value of `r EPV`, which is much lower than the recommended value in the literature.

Once the problem has been demonstrated, we will proceed to launch the methods that allow us to work with high dimensionality. For this, it will not only be necessary to launch the methods themselves that calculate the survival model, but also a cross validation to estimate the optimal values for each methodology.

## Cross Validation

In order to perform survival analysis with our high-dimensional data, we have implemented a series of methods that utilize techniques such as Cox Elastic Net and partial least squares (PLS). To evaluate the performance of these methods, we have implemented cross-validation, which allows us to estimate the optimal parameters for future predictions based on metrics such as AIC, C-INDEX, and AUC. By default AUC metric is used with the "cenROC" evaluator as it has provided the best results in our tests. However, the multiple metrics could be used changed the weights in their parameters. Furthermore, other evaluators have been implemented ("risksetROC", "survivalROC", "cenROC", "nsROC", "smoothROCtime_C" and "smoothROCtime_I").

In addition, we have included options for normalizing data, filtering variables, and setting the minimum EPV, as well as specific parameters for each method, such as the alpha value for Cox Elastic Net and the number of components for PLS models. Overall, our cross-validation methodology allows us to effectively analyze high-dimensional survival data and optimize our model parameters.

## Cox Elastic Net

Cox Elastic Net is based on the R package "glmnet" for survival analysis. However, the structure of the object and the way the analysis is performed has been updated by using our cross-validation methodology to estimate the optimal parameters for future predictions.

During the cross validation, we will iterate over the different alpha values between 0 and 1 (lasso and ridge regression) and as maximum number of variables we select the dimension of the X matrix itself. However, since the EPV requirement will not be met, the number of variables will be limited to the maximum of 9, as previously determined.

```{r, eval=FALSE, warning=F}
# run cv.coxEN
cv.coxen_res <- cv.coxEN(X = X_train, Y = Y_train, 
                         EN.alpha.list = seq(0,1,0.1),
                         max.variables = ncol(X_train),
                         n_run = 2, k_folds = 10, 
                         x.center = T, x.scale = F, 
                         y.center = F, y.scale = F,
                         remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                         remove_non_significant = F, alpha = 0.05, 
                         w_AIC = 0, w_c.index = 0, w_AUC = 1, times = NULL,
                         MIN_AUC_INCREASE = 0.05, MIN_AUC = 0.8, MIN_COMP_TO_CHECK = 3,
                         pred.attr = "mean", pred.method = "cenROC", fast_mode = F,
                         MIN_EPV = 5, return_models = F,
                         PARALLEL = T, verbose = F, seed = 123)
```

```{r, eval=FALSE}
cv.coxen_res #2.7min.
```

```{r}
coxen_model <- coxEN(X = X_train, Y = Y_train, 
                     EN.alpha = 0, #cv.coxen_res$opt.EN.alpha
                     max.variables = 47, #cv.coxen_res$opt.nvar
                     x.center = T, x.scale = F, 
                     y.center = F, y.scale = F, 
                     remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL, 
                     remove_non_significant = F, alpha = 0.05, 
                     MIN_EPV = 5, returnData = T, verbose = F)
```

```{r}
coxen_model
```
As can be seen in the resulting cox model, some of the selected variables have not been significant. If we wish to calculate a model where all selected variables are significant, we should update the remove_non_significant parameter to TRUE.

```{r}
coxen_model <- coxEN(X = X_train, Y = Y_train, 
                     EN.alpha = 0, #cv.coxen_res$opt.EN.alpha
                     max.variables = 47, #cv.coxen_res$opt.nvar
                     x.center = T, x.scale = F, 
                     y.center = F, y.scale = F, 
                     remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL, 
                     remove_non_significant = T, alpha = 0.05, 
                     MIN_EPV = 5, returnData = T, verbose = F)
```

```{r}
coxen_model
```

## PLS-ICOX

In the same way, we will also perform a cross validation for the PLS-based models, starting with the PLS-ICOX methodology. In this case, the internal construction of the weights of the PLS model for the calculation of the components of the X matrix is based on univariate cox models. After this, we are able to reduce the dimensionality of our data set to ultimately launch a cox model with the latent variables or principal components of the PLS model.

```{r, eval=FALSE, message=F}
# run cv.plsicox
cv.plsicox_res <- cv.plsicox(X = X_train, Y = Y_train,
                             max.ncomp = 10,
                             n_run = 2, k_folds = 10, 
                             x.center = T, x.scale = F, 
                             y.center = F, y.scale = F,
                             remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                             remove_non_significant_models = F, alpha = 0.05, 
                             w_AIC = 0, w_c.index = 0, w_AUC = 1, times = NULL,
                             MIN_AUC_INCREASE = 0.05, MIN_AUC = 0.8, MIN_COMP_TO_CHECK = 3,
                             pred.attr = "mean", pred.method = "cenROC", fast_mode = F,
                             MIN_EPV = 5, return_models = F,
                             PARALLEL = T, verbose = F, seed = 123)
```

```{r, eval=FALSE}
cv.plsicox_res #5.45min.
```

```{r, eval=FALSE, fig.small=T}
# plot cv.plsicox
cv.plsicox_res$plot_AUC
```

We will generate a PLS-ICOX model with a total of two principal components based on the results obtained from the cross validation.

```{r}
plsicox_model <- plsicox(X = X_train, Y = Y_train, 
                         n.comp = 3, #n.comp = cv.plsicox_res$opt.comp
                         x.center = T, x.scale = F,
                         y.center = F, y.scale = F,
                         remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                         MIN_EPV = 5, returnData = T, verbose = F)
```

```{r}
plsicox_model
```

## sPLS-DRCOX

Next, we will perform a cross validation for the sPLS-DRCOX methodology. In this case, an sPLS model is run using the deviance residuals of a null Cox model as the Y matrix. The penalties for variable selection follow the strategy used in the R package "plsRcox" from which the idea for the methodology was derived.

```{r, eval=FALSE, message=F}
# run cv.splsdrcox
cv.splsdrcox_res <- cv.splsdrcox(X = X_train, Y = Y_train, 
                                 max.ncomp = 10, eta.list = seq(0,0.9,0.25), #penalty
                                 n_run = 2, k_folds = 10, 
                                 x.center = T, x.scale = F, 
                                 y.center = F, y.scale = F,
                                 remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                                 remove_non_significant_models = F, alpha = 0.05, 
                                 w_AIC = 0,  w_c.index = 0, w_AUC = 1, times = NULL,
                                 MIN_AUC_INCREASE = 0.05, MIN_AUC = 0.8, MIN_COMP_TO_CHECK = 3,
                                 pred.attr = "mean", pred.method = "cenROC", fast_mode = F,
                                 MIN_EPV = 5, return_models = F,
                                 PARALLEL = T, verbose = F, seed = 123)
```

```{r, eval=FALSE}
cv.splsdrcox_res #10.4min
```

```{r, eval=FALSE, fig.small=T}
# plot cv.plsicox
cv.splsdrcox_res$plot_AUC
```

Based on the results obtained through cross validation, we will create a PLS-DRCOX model with a single component and no eta penalty as seen in the cross validation.

```{r}
splsdrcox_model <- splsdrcox(X = X_train, Y = Y_train, 
                             n.comp = 1, eta = 0.75, #n.comp = cv.splsdrcox_res$opt.comp, eta = cv.splsdrcox_res$opt.eta
                             x.center = T, x.scale = F,
                             y.center = F, y.scale = F,
                             remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                             MIN_EPV = 5, returnData = T, verbose = F)
```

```{r}
splsdrcox_model
```

## sPLS-DRCOX MixOmics Penalty

For sPLS-DRCOX methodology we implemented another kind of penalty based on an heuristic variable selection and the MixOmics algorithms. In this case, the penalty is based on a vector of variables. We can select a specific value to select that number of variables, but if we keep the value in NULL, the number of variable will be selected automatically. 

With this method, the user can specify the minimum and maximum number of variables and the number of cutpoints (how many number of variables to test between the minimum and the maximum number of variables) to be tested. After the first iteration, the algorithm will select the optimal number of variables and will further investigate better results between the existing cut points and the optimal value selected.

```{r, eval=FALSE}
# run cv.splsdrcox
cv.splsdrcox_mo_res <- cv.splsdrcox_mixOmics(X = X_train, Y = Y_train, 
                                             max.ncomp = 10, vector = NULL, 
                                             MIN_NVAR = 10, MAX_NVAR = 1000, n.cut_points = 10, EVAL_METHOD = "cenROC",
                                             n_run = 2, k_folds = 10, 
                                             x.center = T, x.scale = F, 
                                             y.center = F, y.scale = F,
                                             remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                                             remove_non_significant_models = F, alpha = 0.05, 
                                             w_AIC = 0,  w_c.index = 0, w_AUC = 1, times = NULL,
                                             MIN_AUC_INCREASE = 0.05, MIN_AUC = 0.8, MIN_COMP_TO_CHECK = 3,
                                             pred.attr = "mean", pred.method = "cenROC", fast_mode = F,
                                             MIN_EPV = 5, return_models = F,
                                             PARALLEL = T, verbose = F, seed = 123)
```

```{r, eval=FALSE}
cv.splsdrcox_mo_res #3.45mins
```

After the cross validation, the user can select the exact number of components and variables to use in their model.

```{r}
splsdrcox_mo_model <- splsdrcox_mixOmics(X = X_train, Y = Y_train, 
                                         n.comp = 2, vector = 534,
                                         x.center = T, x.scale = F,
                                         y.center = F, y.scale = F,
                                         remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                                         MIN_NVAR = 10, MAX_NVAR = 1000, n.cut_points = 5,
                                         MIN_AUC_INCREASE = 0.01,
                                         EVAL_METHOD = "AUC", pred.method = "cenROC", max.iter = 200,
                                         MIN_EPV = 5, returnData = T, verbose = F)

splsdrcox_mo_model
```

## sPLS-DACOX

Finally, we will launch a COX-based sPLS-DA model. This algorithm is the simplest of all and with the least statistical development, but depending on the data set it can provide better results than the previous methods. In this case, we launch an sPLS-DA on the classification of patients according to whether they have suffered or not the study event without taking into account the time until the event. Then, we launch a Cox model using the latent variables of the model and using the entire Y matrix with its times and events/censored.

```{r, eval=FALSE}
# run cv.splsdrcox
cv.splsdacox_res <- cv.splsdacox_mixOmics(X = X_train, Y = Y_train, 
                                          max.ncomp = 10,  vector = NULL,
                                          MIN_NVAR = 10, MAX_NVAR = 1000, n.cut_points = 10, EVAL_METHOD = "cenROC",
                                          n_run = 2, k_folds = 10, 
                                          x.center = T, x.scale = F, 
                                          y.center = F, y.scale = F,
                                          remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                                          remove_non_significant_models = F, alpha = 0.05, 
                                          w_AIC = 0,  w_c.index = 0, w_AUC = 1, times = NULL,
                                          MIN_AUC_INCREASE = 0.05, MIN_AUC = 0.8, MIN_COMP_TO_CHECK = 3,
                                          pred.attr = "mean", pred.method = "cenROC", fast_mode = F,
                                          MIN_EPV = 5, return_models = F,
                                          PARALLEL = T, verbose = F, seed = 123)
```

```{r, eval=FALSE}
cv.splsdacox_res #4min
```

Once the best model is obtained through cross validation, it must be calculated using the desired parameters.

```{r}
splsdacox_mo_model <- splsdacox_mixOmics(X = X_train, Y = Y_train, 
                                         n.comp = 6, vector = 184,
                                         x.center = T, x.scale = F,
                                         y.center = F, y.scale = F,
                                         remove_near_zero_variance = T, remove_zero_variance = F, toKeep.zv = NULL,
                                         MIN_NVAR = 10, MAX_NVAR = 1000, n.cut_points = 5,
                                         MIN_AUC_INCREASE = 0.01,
                                         EVAL_METHOD = "AUC", pred.method = "cenROC", max.iter = 200,
                                         MIN_EPV = 5, returnData = T, verbose = F)

splsdacox_mo_model
```
# Comparing multiple survival models

Next, we will analyze the results obtained from the multiple models to see which one obtains the best predictions based on our data. To do this, we will use the test set that has not been used for the training of any model.

## Comparing for multiple evaluators at the same time.

Initially, we will compare the area under the curve (AUC) for each of the methods according to the evaluator we want. The function is developed to simultaneously evaluate multiple evaluators. However, we will continue working with a single evaluator. In this case "cenROC". On the other hand, we must provide a list of the different models as well as the X and Y test set we want to evaluate.

When evaluating survival model results, we must indicate at which temporal points we want to perform the evaluation. As we already specified a NULL value for the "times" variable in the cross-validation, we are going to let the algorithm to compute them again 

```{r}
lst_models <- list("COX-EN" = coxen_model,
                   "PLS-ICOX" = plsicox_model,
                   "sPLS-DRCOX" = splsdrcox_model,
                   "sPLS-DRCOX-MixOmics" = splsdrcox_mo_model,
                   "sPLS-DACOX-MixOmics" = splsdacox_mo_model)

eval_results <- eval_models4.0(lst_models = lst_models,
                               X_test = X_test, Y_test = Y_test, 
                               pred.method = "cenROC",
                               pred.attr = "mean",
                               times = NULL, max_time_points = 15, 
                               PARALLEL = T)

# lst_evaluators <- c(cenROC = "cenROC", 
#                     risksetROC = "risksetROC")
# 
# eval_results <- purrr::map(lst_evaluators, ~eval_models4.0(lst_models = lst_models,
#                                                            X_test = X_test, Y_test = Y_test, 
#                                                            pred.method = .,
#                                                            pred.attr = "mean",
#                                                            times = seq(1,4,0.5), max_time_points = 15, 
#                                                            PARALLEL = T))
```

We can print the results obtained in the console where we can see, for each of the selected methods, the training time and the time it took to be evaluated, as well as their AIC, C-Index and AUC metrics and at which time points it was evaluated.

```{r}
eval_results
#eval_results$cenROC
```

## Plot comparison

However, we can also obtain graphical results where we can compare each method over time, as well as their average scores using the function "plot_evaluation".

```{r}
lst_eval_results <- plot_evaluation(eval_results)
#lst_eval_results <- plot_evaluation.list(eval_results)
```

After performing the cross-validation, we obtain a list in R that contains two new lists. The first of these refers to the evaluation over time for each of the methods used, as well as a variant where the average value of each of them is shown. On the other hand, we can compare the mean results of each method using: T-test, Wilcoxon-test, anova or Kruskal-Wallis.

```{r, fig.small=T}
lst_eval_results$lst_plots$lineplot
lst_eval_results$lst_plot_comparisons$anova

# lst_eval_results$cenROC$lst_plots$lineplot.mean
# lst_eval_results$cenROC$lst_plot_comparisons$t.test
```

## Computing time comparison (we use the cross validation models)

Another possible comparison is related to the computation times for cross-validation, model creation, and evaluation. In this case, as the cross-validations were not launched for the vignette, we will check which method is the fastest among the others based solely on its construction given the correct parameters.

```{r}
lst_models_time <- list(coxen_model,
                        plsicox_model,
                        splsdrcox_model,
                        splsdrcox_mo_model,
                        splsdacox_mo_model)
```

```{r}
ggp_time <- plot_time.list(lst_models_time)
```

```{r, fig.small=T}
ggp_time
```

# Individual interpretations of the results

Following the cross validation, we have selected the sPLS-DACOX methodology as the most suitable model for our data. We will now study and interpret its results based on the study variables or latent variables used. In this case, we will examine some graphs of the model.

## Forest plots

A forest plot can be obtained as the first graph using the survminer R package. However, the function has been restructured to allow for simultaneous launch of an HDcox class model or a list of HDcox models using the plot_forest() or plot_forest.list() function.

```{r}
lst_forest_plot <- plot_forest.list(lst_models)
```

```{r, fig.small=T}
lst_forest_plot$`sPLS-DRCOX-MixOmics`
```

## PH Assumption

The following graph is related to one of the assumptions of the Cox models, called proportional hazard. 

In a Cox proportional hazards model, the proportional hazards assumption states that the hazard ratio (the risk of experiencing the event of interest) is constant over time for a given set of predictor variables. This means that the effect of the predictors on the hazard ratio does not change over time. This assumption is important because it allows for the interpretation of the model's coefficients as measures of the effect of the predictors on the hazard ratio. Violations of the proportional hazards assumption can occur if the effect of the predictors on the hazard ratio changes over time or if there is an interaction between the predictors and time. In these cases, the coefficients of the model may not accurately reflect the effect of the predictors on the hazard ratio and the results of the model may not be reliable.

In this way, to visualize and check if the assumption is violated, the function plot_proportionalHazard.list() or plot_proportionalHazard() can be called, depending on whether a list of models or a specific model is to be evaluated.

```{r}
lst_ph_ggplot <- plot_proportionalHazard.list(lst_models)
```

Variables or components with a significant P-Value indicate that the assumption is being violated.

```{r, fig.small=T}
lst_ph_ggplot$`sPLS-DRCOX-MixOmics`
```

## Density plots

Another type of graph implemented for all models, whether they belong to the classical branch or to PLS-based models, is the visualization of observations by event according to the values predicted by the Cox models. 

The R package "coxph" allows for several types of predictions to be made on a Cox model that we use in our function, which are:

* Linear predictors "lp": are the expected values of the response variable (in this case, time until the event of interest) for each observation, based on the Cox model. These values can be calculated from the mean of the predictor variable values and the constant term of the model.

* Risk of experiencing an event "risk": is a measure of the probability that an event will occur for each observation, based on the Cox model. The risk value can be calculated from the predictor values and the constant term of the model.

* Number of events expected to be experienced over time with these specific individual characteristics "expected": are the expected number of events that would occur for each observation, based on the Cox model and a specified period of time.

* Terms: are the variables included in the Cox model.

* Survival probability "survival": is the probability that an individual will not experience the event of interest during a specified period of time, based on the Cox model. The survival probability can be calculated from the predictor values and the constant term of the model.

According to the predicted value, we can classify the observations along their possible values and see their distribution for each of the different models.

```{r}
density.plots.lp <- plot_cox.event.list(lst_models, type = "lp")
```

```{r, fig.small=T}
density.plots.lp$`sPLS-DRCOX-MixOmics`$plot.density
density.plots.lp$`sPLS-DRCOX-MixOmics`$plot.histogram
```

## Studying PLS model

For those models based on PLS components, the PLS could be studied in terms of loadings/scores. In order to get the plots, the function plot_PLS_HDcox() has been developed where the user could specify to see "scores", "loadings" or a "biplot" for a couple of latent variables.

```{r}
ggp_scores <- plot_PLS_HDcox(model = lst_models$`sPLS-DRCOX-MixOmics`, 
                             comp = c(1,2), mode = "scores")
```

```{r, fig.small=T}
ggp_scores$plot
```

```{r}
ggp_loadings <- plot_PLS_HDcox(model = lst_models$`sPLS-DRCOX-MixOmics`, 
                               comp = c(1,2), mode = "loadings",
                               top = 10) #length from 0,0
```

```{r, fig.small=T}
ggp_loadings$plot
```

```{r}
ggp_biplot <- plot_PLS_HDcox(model = lst_models$`sPLS-DRCOX-MixOmics`, 
                             comp = c(1,2), mode = "biplot",
                             top = 15,
                             only_top = T,
                             overlaps = 20)
```

```{r, fig.small=T}
ggp_biplot$plot
```

# Understanding the results in terms of the original variables (PLS models)

When a PLS-COX model is computed, the final survival model is based on the PLS latent variables. Although those new variables can explain the survival, in order to understand the disease, new coefficients could be computed in terms of the original variables. 

The idea is to transfer the beta coefficient in the final cox model to an original variable by using its weight in that latent variable. More information could be found in ...!!!!

## Psudobeta

In order to improve the interpretability of a PLS model, a subset of the most influential variables can be selected. In this example, the top 20 variables are chosen. Additionally, non-significant PLS components are excluded by setting the "onlySig" parameter to "TRUE".

```{r}
ggp.simulated_beta <- plot_pseudobeta.list(lst_models = lst_models, 
                                           error.bar = T, onlySig = T, alpha = 0.05, 
                                           zero.rm = T, auto.limits = T, top = 20,
                                           show_percentage = T, size_percentage = 3, verbose = F)
```

The sPSL-DRCOX-MixOmics model was computed using a total of `r lst_models[["sPLS-DRCOX-MixOmics"]]$n.comp` components. Although these components were classified as dangerous for the observations (based on coefficients greater than one), certain variables within the components may still have a protective effect, depending on their individual weights.

The following plot illustrates the pseudo-beta coefficient for the original variables in the sPSL-DRCOX-MixOmics model. As only the top 20 variables are shown, the plot represents only 25.07% of the total linear predictor total value. To view the complete model, all variables would need to be included in the plot by assigning the value NULL to the "top" parameter.

```{r, fig.small=T}
ggp.simulated_beta$`sPLS-DRCOX-MixOmics`$plot
```

## Kaplan-Meier

For a more intuitive understanding of the model, the user can also employ the "getAutoKM.list()" or "getAutoKM()" functions to generate Kaplan-Meier curves. These functions allow the user to view the KM curve for the entire model, a specific component of a PLS model, or for individual variables.

### Full model

To run the full Kaplan-Meier model, the "type" parameter must be set to "LP" (linear predictors). This means that the linear predictor value for each patient will be used to divide the groups (in this case, the score value multiplied by the Cox coefficient). The "surv_cutpoint" function from the R package "survminer" is used to determine the optimal cut-point. Other parameters are not used in this specific method.

```{r}
LST_KM_RES_LP <- getAutoKM.list(type = "LP",
                                lst_models = lst_models,
                                comp = 1:4,
                                top = 10,
                                ori_data = T,
                                BREAKTIME = NULL,
                                only_sig = T, alpha = 0.05)
```

As a result, the Kaplan-Meier curve could be plotted.  

```{r, fig.small=T}
LST_KM_RES_LP$`sPLS-DRCOX-MixOmics`$LST_PLOTS$LP
```

```{r}



lst_ggp <- getTestKM.list(lst_models = lst_models, 
                          X_test = X_test, Y_test = Y_test, 
                          BREAKTIME = NULL, n.breaks = 20,
                          lst_cutoff = 0)

lst_cutoff <- getCutoffAutoKM.list(LST_KM_RES_LP)

lst_ggp <- getTestKM.list(lst_models = lst_models, 
                          X_test = X_test, Y_test = Y_test, 
                          BREAKTIME = NULL, n.breaks = 20,
                          lst_cutoff = lst_cutoff)

ggp <- getTestKM(model = lst_models$`sPLS-DRCOX-MixOmics`, 
                 X_test = X_test, Y_test = Y_test, 
                 BREAKTIME = NULL, n.breaks = 20,
                 cutoff = LST_KM_RES_LP$`sPLS-DRCOX-MixOmics`$info_logrank_num$df_nvar_lrtest$Cutoff)
```

### Components

To generate a Kaplan-Meier curve for a specific component, the "type" parameter must be set to "COMP" (component). This means that the linear predictor is computed using only one component at a time to split the groups. In this case, the "comp" parameter can be used to specify which component should be computed (if multiple components, each one will be computed separately).

```{r}
LST_KM_RES_COMP <- getAutoKM.list(type = "COMP",
                                  lst_models = lst_models,
                                  comp = 1:4,
                                  top = 10,
                                  ori_data = T,
                                  BREAKTIME = NULL,
                                  n.breaks = 20,
                                  only_sig = T, alpha = 0.05)
```

```{r, fig.small=T}
LST_KM_RES_COMP$`sPLS-DRCOX-MixOmics`$LST_PLOTS$comp_1
LST_KM_RES_COMP$`sPLS-DRCOX-MixOmics`$LST_PLOTS$comp_2
```

### Original variables

To generate a Kaplan-Meier curve for original variables, the "type" parameter must be set to "VAR" (variable). In this case, the "ori_data" parameter can be used to determine whether the original or normalized values of the variables should be used.

Additionally, the "top" parameter can be used to plot a specific number of variables, sorted by P-Value (Log-Rank test). The best cutpoint is determined by the "surv_cutpoint" function. Both qualitative and quantitative variables are supported.

```{r}
LST_KM_RES_VAR <- getAutoKM.list(type = "VAR",
                                 lst_models = lst_models,
                                 comp = 1:10, #select how many components you want to compute for the pseudo beta
                                 top = 10,
                                 ori_data = T,
                                 BREAKTIME = NULL,
                                 only_sig = T, alpha = 0.05)
```

```{r, fig.small=T}
LST_KM_RES_VAR$`sPLS-DRCOX-MixOmics`$LST_PLOTS$hsa_miR_222
LST_KM_RES_VAR$`sPLS-DRCOX-MixOmics`$LST_PLOTS$hsa_miR_182
```

# New patients

If new patients are to be studied using a trained model, several functions have been implemented.

To demonstrate, an observation from a test dataset will be used.

```{r}
new_pat <- X_test[1,,drop=F]
```

As shown, this is a patient who did not reach the event, and the last time the patient was checked before being censored was at time 226.

```{r}
knitr::kable(Y_test[rownames(new_pat),])
```

The function "plot_pseudobeta_newPatient.list()" or "plot_pseudobeta_newPatient()" allows the user to compare the characteristics of a new patient with the pseudo-beta values obtained from a specific model. The goal is to understand in a general way which variables are associated with an increased risk of an event or a decreased risk, in comparison to the variables predicted by the model.

```{r}
ggp.simulated_beta_newPat <- plot_pseudobeta_newPatient.list(lst_models = lst_models, 
                                                             new_pat = new_pat,
                                                             error.bar = T, onlySig = T, alpha = 0.05,
                                                             zero.rm = T, auto.limits = T, show.betas = T, top = 20)

# ggp.simulated_beta_newPat <- plot_pseudobeta_newPatient(model = lst_models$`sPLS-DRCOX-MixOmics`, 
#                                                         new_pat = new_pat,
#                                                         error.bar = T, onlySig = T, alpha = 0.05,
#                                                         zero.rm = T, auto.limits = T, show.betas = T, top = 20)
```
On the left, a linear predictor value is shown for the observation and each variable. On the right, the pseudo-beta coefficients for the model's original variables are illustrated. This allows the user to compare the direction of the linear predictors per variable. A change in direction means that the variable's value is below the mean for hazard variables or above the mean for protective ones. Having an opposite direction for hazard variables and maintaining the direction for protective variables makes the observation safer over time.

```{r, fig.small=T}
ggp.simulated_beta_newPat$`sPLS-DRCOX-MixOmics`$plot
```

## Add patient to density plot

Additionally, the new observation can be added to the density and histogram plots that the model has computed. While this function can be useful in visualizing the results, it is important to note that no definitive conclusions can be drawn from these types of plots alone. They serve as an additional means of viewing the predictive values for the new patient ("lp", "risk", "expected", "terms", "survival") in relation to the training dataset and its histogram or density plot, along with the events or censored patients.

```{r}
pat_density <- plot_patient.eventDensity(patient = new_pat, 
                                         time = NULL, 
                                         model = lst_models$`sPLS-DRCOX-MixOmics`, 
                                         type = "lp")
```

```{r, fig.small=T}
pat_density
```

```{r}
pat_histogram <- plot_patient.eventHistogram(patient = new_pat, 
                                             time = NULL, 
                                             model = lst_models$`sPLS-DRCOX-MixOmics`, 
                                             type = "lp")
```

```{r, fig.small=T}
pat_histogram
```

```{r, eval=F}
#plot_divergent.biplot - for num and qual variables
```

## COX compare multiple patients

Furthermore, similar plots can be generated for multiple patients at the same time. For example, by selecting 5 patients, individual linear predictors for each variable and the final linear predictor using all the variables used in the survival model can be plotted.

```{r}
knitr::kable(Y_test[1:5,])
```

To generate the figure, the function "plot_LP.multiplePatients.list()" or "plot_LP.multiplePatients()" can be used.

```{r}
lst_cox.comparison <- plot_LP.multiplePatients.list(lst_models = lst_models, 
                                                   df.pat = X_test[1:5,], 
                                                   error.bar = T, zero.rm = T, onlySig = T, alpha = 0.05, top = 5)
```

```{r, fig.small=T}
lst_cox.comparison$`sPLS-DRCOX-MixOmics`$plot
```
